---
title: "Calculus"
editor: visual
code-fold: true
fig-width: 5
fig-height: 3
---

```{r}
#| code-summary: Packages
library(ggplot2)
theme_set(
  theme_minimal(base_family = "Amiri") +
  theme(axis.title.y = element_text(angle = 0, vjust = 1/2))
)
```

## Derivatives

**Limit Definition**

$$
\begin{align}
f'(x) = \lim_{h \to 0} \frac{f(x + h) - f(x)}{h}, &&\text{provided the limit exists}
\end{align}
$$ {#eq-deriv-limit}

Derivatives are used to understand instantaneous rates of change, i.e., when $\Delta x \approx 0$.

This formula is equivalent:

$$
f^\prime (a) = \lim_{b \to a} \frac{f(b) - f(a)}{b-a}
$$

For example, the area of a growing square is $A(x) = x^2$. How does $A(x)$ change every time $x$ changes? We have that, for any given value of $x$, the rate of instantaneous change is given by:

$$
\begin{align}
A^\prime (x) = 2x && \text{or} && \frac{dA}{dx} = 2x
\end{align}
$$

*Note. When* $\Delta$ *is infinitely small we change notation to* $d$*.*

If we zoom in on *any* function at a given point $a$, a tangent line at $a$ will be given by the following formula:

$$
y = f(a) + f^\prime (a) (x-a)
$$

<aside>*linear approximation of* $f$ *near* $x=a$</aside>

**Critical points**

$c$ is a critical point for the smooth function $f$ iff $f^\prime(c) = 0$.

In order to figure out if $c$ is a local maximum or local minimum, we use the *second derivative.* The second derivative of $f$ measures the rate of change of the first derivative. In a nutshell, we find critical points by solving $f^\prime = 0$. We then plug these values into $f^{\prime\prime}$. If the result is positive, the critical point is a local minimum; if the result is negative, the critical point is a local maximum.

*This procedure does not always work.*

*Note. For higher order derivatives we use the* $f^{(n)}(x)$ *notation, where* $f^{(0)}(x) = f(x)$*. We can also use the symbol* $\frac{d^nf}{dx^n}$ *for this purpose.*

**Derivative Rules:**

*The constant rule*---i.e., constants can move into and out of derivatives.

$$
\frac{d}{dt} \bigg[c f(t) \bigg] = c \frac{d}{dt} \bigg[f(t)\bigg] = c \cdot f^\prime (t)
$$ {#eq-constant-rule}

*The sum rule---*i.e., the derivative of the sum is the sum of the derivatives.

$$
\frac{d}{dt}\bigg[ f(t) + g(t) \bigg] = f^\prime (t) + g^\prime (t)
$$ {#eq-sum-rule}

*The product rule.*

$$
\frac{d}{dx} \bigg[f(x) g(x) \bigg] = f(x) g^\prime (x) + g(x) f^\prime(x)
$$ {#eq-product-rule}

*The power rule.*

$$
\frac{d}{dx} \bigg[ x^n\bigg] = n \cdot x^{n-1}
$$ {#eq-power-rule}

*Note. All these rules make it possible to estimate the derivative of any polynomial of the form* $c_nx^n + c_{n-1} x^{n-1} + \dots + c_1$*.*

*The chain rule* allows us to find derivatives of compositions.

$$
\frac{d}{dx} \bigg[ (f \circ g)(x) \bigg] = f^\prime (g(x)) \cdot g^\prime (x)
$$ {#eq-chain-rule}

<aside>$$
(f \circ g)(x) = f(g(x))
$$</aside>

We can use @eq-power-rule and @eq-chain-rule to figure out the following:

$$
\frac{d}{dx} \bigg[ \frac{1}{g(x)} \bigg] = - \frac{g^\prime (x)}{g(x)^2}
$$

Using this with @eq-product-rule, we can figure out the *quotient rule:*

$$
\frac{d}{dx} \bigg[ \frac{f(x)}{g(x)}\bigg] = f^\prime (x) \Bigg( \frac{1}{g(x)} \Bigg) + f(x) \Bigg(- \frac{g^\prime (x)}{g(x)^2} \Bigg) = \frac{f^\prime (x) g(x) - f(x) g^\prime (x)}{g(x)^2}
$$ {#eq-quotient-rule}

## Integrals

What is the area $A(a, b)$ below the graph $y = f(x) \geq 0$, above the $x$-axis, and between $x=a$, and $x=b$? We can estimate $A(a, b)$ with a sum, and then argue that the estimate becomes exact as the number of terms approximates $\infty$. This is called a *Riemann sum.*

$$
A(a, b) = \underbrace{\int_a^b \overbrace{f(x)}^{\small \text{integrand}} dx}_\text{integral}
$$

The elongated "s" stands for "sum," when the number of terms to be summed approaches infinity. The decorations on the "s" are called the *lower* and *upper limits of integration*.

The symbols $f(x) \cdot dx$ make it clear that the terms of a Riemann sum are the areas of rectangles:

$$
\begin{align}
\text{base: } (x_{n+1} - x_n), &&\text{height: }f \bigg( \frac{x_n + x_{n+1}}{2} \bigg)
\end{align}
$$

For example,

$$
\int_0^1 x^2 dx = \frac{x^3}{3} \Bigg|_0^1 = \frac{1^3}{3} - \frac{0^3}{3} = \frac{1}{3}
$$

represents the area shaded in pink in the following graph:

```{r}
ggplot() + 
  xlim(-0.5, 1.5) + 
  geom_function(fun = \(x) x^2) +
  stat_function(geom = "area", fun = \(x) x^2, xlim = c(0, 1), fill = "pink") + 
  labs(x = "x")
```

```{r}
#| code-fold: show
integrate(\(x) x^2, lower = 0, upper = 1)

n <- 100
x <- seq(0, 1, length.out = n)

result <- 0
for (i in 1:(n - 1)) {
  result <- result + (x[i + 1] - x[i]) * ((x[i] + x[i + 1])/2)^2
}

result
```

<aside>100 is much smaller than $\infty$</aside>

We use the area under the curve in this graph to think about many things---e.g., displacement of an object with $x$ as time and $y$ as velocity; probability distributions of arbitrary random variables with $y$ as probability densities; etc.

**The Fundamental Theorem of Calculus**

$$
f(b) - f(a) = \int_a^b f^\prime (x) dx
$$ {#eq-fundamental-theorem}

We have already seen an example of this when looking for shaded area under $f(x) = x^2$ between $0$ and $1$ in the previous example:

$$
\int_0^1 x^2 dx = \int_0^1 \frac{d}{dx} \bigg[ \frac{1}{3} x^3\bigg] dx = \bigg(\frac{1}{3}x^3\bigg) \Bigg|_0^1 = \frac{1}{3} (1^3) - \frac{1}{3} (0^3) = \frac{1}{3}
$$

<aside>$f|_a^b$ is shorthand for $f(b) - f(a)$</aside>

**Anti-derivatives**

If $F^\prime (x) = f(x)$, then the definite integral of $f$ over $[a,b]$ is

$$
\int_a^b f(x) dx = F(b) - F(a)
$$

The function $F(x)$ is called *an* anti-derivative of $f(x)$.

All anti-derivatives of $f(x)$ are given by $F(x) + C$ for any *constant of integration* $C$.

*The Power Rule*

$$
\begin{align}
\int x^n dx = \frac{1}{n+1} x^{n+1} + C, &&\text{for } n \neq -1
\end{align}
$$

and for $n=1$ we have

$$
\int x^{-1} dx = \ln |x| + C
$$

*Note. We use "*$\int$" for anti-derivatives because they are closely related to integrals thanks to the fundamental theorem of calculus shown in @eq-fundamental-theorem.

*The Sum Rule*

$$
\int \bigg[ f(x) + g(x) \bigg] dx = \int f(x) dx + \int g(x) dx
$$

*The Constant Rule*

$$
\int \bigg[ c f(x) \bigg] dx = c \int f(x) dx
$$

## Infinite Sums

*Preliminary remarks*

A **sequence** $A_n$ is an ordered list of numbers where the *index* $n$ is drawn from a fixed subset of the integers. In general, the sequence $A_n$ has a limit $L$ if we can make $|A_n - L|$ as small as we like by picking $n$ large enough.

$$
\lim_{n \to \infty} A_n= L
$$

Example: *logistic sequence.*

Suppose a population $p_n$ that increases in size over time relative to some carrying capacity (thus, $p$ is a proportion).

$$
p_{n+1} = c \times p_n \times (1 - p_n)
$$

<aside>recursive equation</aside>

Here, $c$ is a positive constant. If the population growth given by this formula eventually stabilizes, we can think of the *long-term population* as $\lim_{n \to \infty} p_n$, assuming that this limit actually exists.

$$
L = \frac{c-1}{c}
$$

<aside>

Plug in $L$ for both $p_n$ and $p_{n+1}$ and solve for $L$.

</aside>

*However, we don't know if the* $\lim_{n \to \infty} p_n$ *actually exists for all values of* $c$ *and* $p_o$ *(it doesn't).*

**Infinite sums**

An infinite sum is the limit of the sequence of partial sums:

$$
\sum_{j = 1}^\infty a_j = \lim_{n \to \infty} \sum_{j = 1}^n a_j
$$ {#eq-infinite-sum}

When this limit exists, we say that the sum **converges**; otherwise we say that it **diverges.**

Example: *the geometric sum*

$$
\sum_{j=0}^\infty r^j
$$

Geometric sums have a special place in calculus for a variety of reasons, the most important being that we know precisely when they converge and what they converge to.

It's useful to think of the following sequences:

$$
\begin{align}
S_n &= 1 + r + r^2 + \dots + r^n, \\
S_{n+1} &= 1 + r + r^2 + \dots r^{n+1}
\end{align}
$$

From which we can establish that:

$$
\begin{align}
S_{n+1} &= r S_n + 1, \\
S_{n+1} &= S_n + r^{n+1}
\end{align}
$$

<aside>

Multiply $S_n$ by $r$ and add $1$ to get the first equation.

</aside>

As it turns out, the geometric sum converges if $|r| < 1$ and diverges if $|r| \geq 1$.

Assuming that the limit exists, and thus $|r| < 1$, we can get the following:

$$
\lim_{j \to \infty} \sum_{j=0}^j r^j = \frac{1}{1-r}
$$

<aside>

Remember to plug in $L$ for both $S_{n+1}$ and $S_n$.

</aside>

**Taylor Series**

A smooth function $f$ can be approximated near any given point $x=a$ by

$$
y = \underbrace{f(a) + f^{(1)} (a) (x-a) + \frac{1}{2!}f^{(2)}(a)(x-a)^2}_{\text{quadratic approximation}} + \frac{1}{3!}f^{(3)}(a)(x-a)^3 + \dots
$$ {#eq-taylor-series}

## Other

Euler's number $e$ is defined in terms of a limit approaching infinity.

$$
e = \lim_{x \to \infty} \bigg( 1 + \frac{1}{x} \bigg)^x \approx 2.718282...
$$

We also call this a *horizontal asymptote.*

```{r}
foo <- function(x) (1 + 1/x)^x

ggplot() + 
  xlim(-5, 50) +
  coord_cartesian(ylim = c(0.5, 5)) +
  labs(x = "x", y = "f(x)") +
  geom_function(fun = foo,  n = 1e3) + 
  geom_hline(yintercept = exp(1), linetype = "dashed")
```

Note that the limit is undefined as $x \to 0$.
